### 2017-01-25 manuscript.rmd
more polishing

The KMRR study considered several publicly available datasets. Here we consider their conclusions in the context of one of these studies, Brawand *et al.* 2011 [@Brawand:2011du]. This dataset of gene expression for six organs across 10 species (nine mammals and a bird) is the best sampled in their analyses. Their expression summary statistic is Tau [@Yanai:2005dh, @KryuchkovaMostacci:2016in], an indicator of tissue specificity of gene expression. It can range from a value of 0, which indicates no specificity (*i.e.*, uniform expression across tissues), to a value of 1, which indicates high specificity (*i.e.*, expression in only one tissue). It is convenient in that it is a single number of defined range for each gene, though of course since the original expression is multidimensional this means much information is discarded. For example, it does not convey which tissue expression is specific to. Homologous genes could each have a tau value of 1, but one gene could have expression that is specific to the kidney and the other to the brain, for example.

KMRR's published analyses are based on pairwise comparisons (Figure 1a) between tau within each gene family. They found the correlation of tau for orthologs to be significantly greater than the correlation of tau for paralogs, *i.e.* that orthologs tend to be more similar to each other than paralogs. From this they concluded that their analyses support the ortholog conjecture. They also went one step further. They concluded that this pattern provides support for a particular evolutionary process, that "tissue-specificity evolves very slowly in the absence of duplication, while immediately after duplication the new gene copy differs".

@@

We reanalyzed the the KMRR study using phylogenetic comparative methods, focusing on the dataset of Brawand *et al.* 2011 [Brawand:2011du]. For each internal node in each gene tree (Figure 1b), we measured the phylogenetically independent contrast (PIC) in the expression across this node [@Felsenstein:1985ua]. This is the difference in values of descendant nodes (*i.e.*, the evulutionary change along the branches that descend from the node) scaled by the expected variance (determined largely by the lengths of these branches). These contrasts were then annotated by whether they are made across branches that descend from a speciation or duplication event. The original description of independent contrasts [@Felsenstein:1985ua] focused on assessing covariance between changes in two traits on a tree. Our use of it to look for differences in evolutionary changes of one trait (differential expression) between two categories of nodes (speciation and duplication) on a tree is a bit different than the usual application of the method.

@@

> Figure XXExpectations | (a) Null hypothesis. (b) Under the ortholog conjecture.

To test the ortholog conjecture it is essential to have a null hypothesis that results in a distinct prediction from the ortholog conjecture. A suitable null hypothesis is that there is no difference in evolutionary change in differential expression along branches that descend from speciation or duplication events (Figure 1b). Under this hypothesis, we would predict that contrasts across speciation nodes and duplication nodes are drawn from the same distribution (Figure XXEXPECTATIONS a). Under the alternative hypothesis specified by the ortholog conjecture, that there is a higher rate of change following duplication events than speciation events, we would expect to see the distribution of duplication contrasts shifted to higher values relative to the speciation contrasts (Figure XXEXPECTATIONS b).

When we reanalyze the data with phylogenetic independent contrasts (Fig. 1B), we did not find increased evolutionary change in expression following duplication events (Fig. XX).

@@

@@

		if( remaining > 3 ){
			pruned = treeio::drop.tip( nhx, to_drop )
			return( pruned )
		}
		else{
			return( NA )
		}
	}

@@

Of the `r length(gene_trees)` trees that were parsed, `r length(gene_trees_sub)` passed taxon sampling criteria and `r length(gene_trees_calibrated)` were successfully time calibrated.

@@

There were `r nodes_contrast %>% filter(D=="Y") %>% nrow()` duplication nodes and `r nodes_contrast %>% filter(D=="N") %>% nrow()` speciation nodes.

Given the ortholog conjecture, we would expect to see larger expression contrasts following duplication events than speciation events. Instead,we find no such distribution (Figure XXKMRR). The Wilcox rank test does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (p value = `r wilcox_test_result$p.value`). Our results therefore find no evidence for the ortholog conjecture in this system.

@@

We next examined the possibility that ascertainment biases were differentially impacting the inference of evolutionary change in differential expression following duplication and speciation events. We focused on two possible sources of bias - node depth and branch length.

While the age of speciation nodes is constrained, duplication nodes can be much older and therefore have a wider range of ages (Supplementary Figure XXAscertainment a). Biological reasons for this include the fact that duplication events can be older than the root of the species tree, but speciation events cannot. There are also technical factors that can lead to an excess of duplication events deeper in the tree, including the impact of assembly errors that lead to the overestimation of deep duplications [@Hahn:2007cj]. If independent contrast values also tended to to be lower at greater node depth, it could interact with the preponderance of duplications at greater depth to create a pattern of lower contrasts following duplication events. When we remove duplication nodes that are older than the oldest speciation node, our results are unchanged and we cannot reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (Supplementary Figure XXAscertainment b).

The independent contrast across a node is the amount of change observed between the daughter notes, scaled by the expected variance [Felsenstein:1985ua]. The expected variance is principally determined by the lengths of the branches leading from the node to these daughters. The same observed difference between daughter nodes will lead to a larger contrast the shorter the branches are to these nodes, as this indicates a greater rate of evolutionary change. The expected variance of contrasts for speciation nodes is constrained by the branch lengths on the species tree, but the expected variance of contrasts for duplications can range much wider (Figure XXAscertainment c).

The fact that the expected variance of speciation contrasts is a subset of the range of the expected variance of duplication contrasts could lead to ascertainment biases. We therefore also examine only the contrasts that fell within in the range of expected variance seen for speciation contrasts, excluding duplication contrasts that fall outside of this range. This reanalysis does not reject the ortholog conjecture either (Figure XXAscertainment d).

@@

In order to better understand why our phylogenetic analysis seem to support a different conclusion (*i.e.*, no support for the ortholog conjecture) than the published analysis of KMRR [@KryuchkovaMostacci:2016iw] (*i.e.*, strong support for the ortholog conjecture), we first checked to make sure we could reproduce their specific result. This is important since we are only looking at at a subset of the data they considered, the Brawand *et al.* 2011 [@Brawand:2011du] dataset. We can replicate their result - we find a higher tau correlation between orthologs (R=`r round( ortholog_r, 3) `) then between paralogs (R=`r round( paralog_r, 3)`).

Why is it that pairwise methods and phylogenetic methods lead to the opposite conclusion? The reason is that they describe different things. Pairwise comparisons describe contemporary patterns, while phylogenetic methods infer historical processes. There need not be a different process of evolution following speciation and duplication for there to be different patterns for orthologs and paralogs. That is because there is one other important link between evolutionary process and contemporary patterns - the structure of the tree.

Paralogs could be more different from each other than orthologs if they tend to be more distantly related to each other than orthologs are. That would provide more time for differences to accumulate between them, regardless of the history of duplication and speciation. This is, in fact, what we find in this dataset. While the mean distance between orthologs is `r round( ortholog_distance_mean, 1 )` years, the mean distance between paralogs is `r round( paralog_distance_mean, 1 )` years. This is at least in part because, while the maximum distance between orthologs is two times the depth of the oldest speciation event in the tree, many duplications precede this oldest sampled speciation event. The distance between paralogs resulting from these deeper duplications can be much larger.

If we consider only the paralogs that arise from the duplication events the same age or younger than the oldest speciation event, the observed pattern is very different. The correlation of tau between these shallower paralog pairs, R=`r round( paralog_subset_r, 3)`, is much stronger than for the correlation between all paralog pairs, R=`r round( paralog_r, 3)`. This is much closer to the ortholog comparison, which has R=`r round( ortholog_r, 3) `.

The other issue with the pairwise comparisons is that many are summarizing changes along many branches in the phylogeny, not just the branches that follow the divergence event. Two paralogs that diverged from a duplication event deep in the tree may have many subsequent duplication and speciation events, and changes along all these branches will impact the final pattern. The phylogenetic methods isolate the changes under consideration. There may still be missing speciation events, due to extinction and incomplete taxon sampling, and missing duplication events, due to gene loss. But these affect both methods. The phylogenetic methods avoid diluting the change that occurs along the branches that follow the node in question with changes along all subsequent branches.

The reason that paralogs pairs have tau values that are more different than ortholog pairs is that they tend to be more distantly related, not because there are differences in evolutionary rates of change in tau following duplication and speciation events. This means that having information on whether two genes are orthologs or paralogs provides little added information about expression beyond knowing how distantly related the two genes are. This has several implications. One is that it is an example of the limitations of the terms orthology and paralogy [@Dunn:2016er]. These terms can have straightforward meaning in small gene trees with simple duplication/speciation histories, but the utility of the terms breaks down on larger more complex gene trees. Orthology and paralogy are annotations on the tips of the tree that are derived from the infered structure of the tree and annotations of duplication and speciation at internal tree nodes. In this sense, orthology and paralogy are statements about the internals of the tree that are distilled into statements at the tips of the tree. Much is lost in the process, though, and for most questions it is much more direct to focus on the structure of the tree and the inferred processes of internal nodes. In this case, KMRR [@KryuchkovaMostacci:2016iw] relied on orthology and paralogy annotations that are derived from the infered duplication and speciation history on gene phylogenies [@Herrero:2016jj]. These are the same pre-computed trees we use in our phylogenetic re-analyses of their study.

@@

While we do demonstrate a common problem with pairwise comparisons in the Levin *et al.* analysis, we did not perform a phylogenetic reanalysis of this study, as we did for the KMRR study. This is because the similarity metric computed in the pairwise comparisons of Levin *et al.* are based on different genes for different species pairs. This means that there is not a trait that can be mapped onto the tree, as KMRR's tau can be. A full phylogenetic reanalysis would of course be possible be using upstream analysis products to re-derive a new expression summary statistic that was suitable for mapping onto a phylogenetic tree.

@@

>>>>

The KMRR study considered several publicly available datasets. Here we consider their conclusions in the context of one of these studies, Brawand *et al.* 2011 [@Brawand:2011du]. This dataset of gene expression for six organs across 10 species (nine mammals and one bird) is the best sampled in their analyses. Their expression summary statistic is Tau [@Yanai:2005dh, @KryuchkovaMostacci:2016in], an indicator of tissue specificity of gene expression. It can range from a value of 0, which indicates no specificity (*i.e.*, uniform expression across tissues), to a value of 1, which indicates high specificity (*i.e.*, expression in only one tissue). It is convenient in that it is a single number of defined range for each gene, though of course since the original expression is multidimensional this means much information is discarded. This includes information about which tissue expression is specific to. For example, if one gene has expression specific to the brain and another expression specific to the kidney, both would have a Tau of 1.

KMRR's published analyses are based on pairwise comparisons (Figure 1a) between Tau within each gene family. They found the correlation of Tau for orthologs to be significantly greater than the correlation of Tau for paralogs, *i.e.* that orthologs tend to be more similar to each other than paralogs. From this they concluded that their analyses support the ortholog conjecture. They also concluded that this pattern provides support for a particular evolutionary process, that "tissue-specificity evolves very slowly in the absence of duplication, while immediately after duplication the new gene copy differs".

@@

We reanalyzed the the KMRR study using phylogenetic comparative methods, focusing on the dataset of Brawand *et al.* 2011 [@Brawand:2011du]. For each internal node in each gene tree (Figure 1b), we measured the phylogenetic independent contrast[@Felsenstein:1985ua] (PIC) of the expression statistic Tau. This is the difference in values of Tau for descendant nodes scaled by the expected variance (determined largely by the lengths of these branches). These contrasts were then annotated by whether they are made across a speciation or duplication event. The original description of independent contrasts [@Felsenstein:1985ua] focused on assessing covariance between changes in two traits on a tree. Our use of contrasts is a bit different-- we look for differences in evolutionary changes of one trait (differential expression) between two categories of nodes (speciation and duplication) on a tree.

@@

> Figure XXExpectations | (a) Under the null hypothesis that there is no difference in the rate of evolution after duplication or speciation events, the distribution of PICs would be the same across both types of nodes. (b) Under the ortholog conjecture, contrasts across duplication nodes would tend to be larger than contrasts across speciation nodes.

To test the ortholog conjecture it is essential to have a null hypothesis that makes predictions distinct from the ortholog conjecture. A suitable null hypothesis is that there is no difference in evolutionary change in differential expression along branches that descend from speciation or duplication events (Figure 1b). Under this hypothesis, we would predict that contrasts across speciation nodes and duplication nodes are drawn from the same distribution (Figure XXEXPECTATIONS a). Under the alternative hypothesis specified by the ortholog conjecture, that there is a higher rate of change following duplication events than speciation events, we would expect to see the distribution of duplication contrasts shifted to higher values relative to the speciation contrasts (Figure XXEXPECTATIONS b).

@@

	# The minimum number of genes with expression data that a gene needs to have to be considered
	min_genes_with_expression = 4

@@

		if( remaining >= min_genes_with_expression ){
			pruned = treeio::drop.tip( nhx, to_drop )
			return( pruned )
		}
		else{
			return( NA )
		}
	}

@@

@@

When we reanalyze the data with phylogenetic independent contrasts (Fig. 1B), we did not find increased evolutionary change in expression following duplication events (Fig. XXKMRR). Of the `r length(gene_trees)` trees that were parsed from ENSMBL Compara [@Herrero:2016jj], `r length(gene_trees_sub)` passed taxon sampling criteria (`r min_genes_with_expression` genes) after removing tips without Tau values. Of these, `r length(gene_trees_calibrated)` were successfully time calibrated. These calibrated pruned trees were used to calculate phylogenetic independent contrasts for Tau. They contain `r nodes_contrast %>% filter(D=="Y") %>% nrow()` duplication nodes and `r nodes_contrast %>% filter(D=="N") %>% nrow()` speciation nodes.

Given the ortholog conjecture, we would expect to see larger contrasts following duplication events than speciation events. Instead, we find no such distribution (Figure XXKMRR). The Wilcox rank test does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (p value = `r wilcox_test_result$p.value`). Our results therefore find no evidence for the ortholog conjecture in this system.

@@

We next examined the possibility that ascertainment biases were differentially impacting the inference of evolutionary following duplication and speciation events. We focused on two possible sources of bias - node depth and branch length.

While the age of speciation nodes is constrained, duplication nodes can be much older and therefore have a wider range of ages (Supplementary Figure XXAscertainment a). This is because many gene duplication events are older than the most recent common ancestor of the species in the study. There are also technical factors that can lead to an excess of duplication events deeper in the tree. Assembly errors, for example, tend to lead to the overestimation of deep duplications [@Hahn:2007cj]. If independent contrast values also tended to to be lower at greater node depth, it could interact with the preponderance of duplications at greater depth to create a pattern of lower contrasts associated with duplication events. To test for such an effect, we remove duplication nodes that are older than the oldest speciation node. Our results are unchanged and this reduced dataset does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (Supplementary Figure XXAscertainment b).

The independent contrast across a node is the amount of change observed between the daughter notes, scaled by the expected variance [Felsenstein:1985ua]. The expected variance is principally determined by the lengths of the branches leading from the node to these daughters. The shorter the total length of the two branches leading to dauter nodes, the larger the contrast for a given observed difference. This is because the same difference across a shorter total branch length indicates a greater rate of evolutionary change. The expected variance of contrasts for speciation nodes is constrained by the branch lengths on the species tree, but the expected variance of contrasts for duplications has a much wider range (Figure XXAscertainment c). This could lead to biases if the lengths of branches that descend from duplication nodes tend to be overestimated. We thereforeexamined only the contrasts that fell within in the range of expected variance seen for speciation contrasts, excluding duplication contrasts that fall outside of this range. This reanalysis does not reject the ortholog conjecture either (Figure XXAscertainment d), indicating that branch length bias is not responsible for the result.

@@

In order to better understand why our phylogenetic analysis seems to support a different conclusion (*i.e.*, no support for the ortholog conjecture) than the published analysis of KMRR [@KryuchkovaMostacci:2016iw] (*i.e.*, strong support for the ortholog conjecture), we first checked to make sure we could reproduce their result based on pairwise analyses. This is important since we are only looking at at a subset of the data they considered, the Brawand *et al.* 2011 [@Brawand:2011du] dataset for gene trees that could be successfully time calibrated. We can replicate their result - we find a higher Tau correlation between orthologs (R=`r round( ortholog_r, 3) `) then between paralogs (R=`r round( paralog_r, 3)`).

Why is it that pairwise methods and phylogenetic methods lead to opposite conclusions? The reason is that they describe different things. Pairwise comparisons describe contemporary patterns, while phylogenetic methods infer historical processes. There need not be a different process of evolution following speciation and duplication for paralogs to be more different than orthologs. This could be due to the structure of the gene phylogenies. If paralogs tend to be more distantly related to each other than orthologs, then there would be more time for differences to accumulate even if the rate of change is the same between the two. This is, in fact, the case for these data. While the mean distance between orthologs is `r round( ortholog_distance_mean, 1 )` years, the mean distance between paralogs is `r round( paralog_distance_mean, 1 )` years. As described above, this is because the oldest speciation event is be definition the most recent common ancestor of the the species included in the study, but many gene families underwent duplication before this time.

To test the hypothesis that distantly related paralogs arising from these older duplication events are the reason for the lower correlation of Tau between paralogs than between orthologs, we removed them. When we consider only the paralogs that arise from duplication events the same age or younger than the oldest speciation event, the observed pattern is very different. The correlation of Tau between these shallower paralog pairs, R=`r round( paralog_subset_r, 3)`, is much stronger than for the correlation between all paralog pairs, R=`r round( paralog_r, 3)`. This is much closer to the ortholog correlation, which has R=`r round( ortholog_r, 3) `.

The reason that paralog pairs have Tau values that are more different than ortholog pairs is that they tend to be more distantly related, not because there are different rates of Tau evolution following duplication and speciation events. This means that having information on whether two genes are orthologs or paralogs provides little added information about expression beyond knowing how distantly related the two genes are. This has several implications. It is an example of the limitations of the concepts of orthology and paralogy [@Dunn:2016er]. These terms can have straightforward meaning in small gene trees with simple duplication/speciation histories, but the utility of the terms breaks down on larger more complex gene trees. Orthology and paralogy are annotations on the tips of the tree that are derived from the infered structure of the tree and annotations of duplication and speciation at internal tree nodes. In this sense, orthology and paralogy are statements about the internals of the tree that are distilled into statements at the tips of the tree. Much is lost in the process, though. In this case, KMRR [@KryuchkovaMostacci:2016iw] relied on orthology and paralogy annotations that are derived from the infered duplication and speciation history on gene phylogenies [@Herrero:2016jj]. These are the same pre-computed trees we use in our phylogenetic re-analyses of their study. For most questions it is much more direct to focus on the structure of the tree and the inferred processes within the tree, such as which internal nodes are duplication or speciation events and how much change occurs along the branches.

@@

While we do demonstrate a common problem with pairwise comparisons in the Levin *et al.* analysis, we did not perform a phylogenetic reanalysis of this study, as we did for the KMRR study. This is because the similarity metric computed in the pairwise comparisons of Levin *et al.* are based on different genes for different species pairs. This means that there is not a trait that can be mapped onto the tree, as KMRR's Tau can be. A full phylogenetic reanalysis would of course be possible be using upstream analysis products to re-derive a new expression summary statistic that was suitable for mapping onto a phylogenetic tree.

@@

The other issue with the pairwise comparisons is that, except at the tips, they summarize changes along many branches in the phylogeny. Two paralogs that diverged from a duplication event deep in the tree may have many subsequent duplication and speciation events, and changes along all these branches will impact the final pairwise comparison. Phylogenetic methods have the advantage of isolating the changes under consideration (Fig. 1b). The phylogenetic methods avoid diluting the change that occurs along the branches that follow the node in question with changes along all subsequent branches. There may still be missing speciation events, due to extinction and incomplete taxon sampling, and missing duplication events, due to gene loss, but these omissions affect both methods.
